The role of cognateness in non-native spoken word recognition

### 2021-11-25 Rmd/manuscript.tex
Starts manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The role of cognateness in non-native spoken word recognition},
  pdfauthor={Gonzalo Garcia-Castro1, Serene Siow2, Nuria Sebastian-Galles1, \& Kim Plunkett2},
  pdflang={en-EN},
  pdfkeywords={cognate, word recognition, translation, non-native, spoken word recognition},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Cognateness and non-native word recognition}
\keywords{cognate, word recognition, translation, non-native, spoken word recognition\newline\indent Word count: X}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifXeTeX
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{The role of cognateness in non-native spoken word recognition}
\author{Gonzalo Garcia-Castro\textsuperscript{1}, Serene Siow\textsuperscript{2}, Nuria Sebastian-Galles\textsuperscript{1}, \& Kim Plunkett\textsuperscript{2}}
\date{}

\authornote{

GG and SS contributed equally and share first authorship.

Correspondence concerning this article should be addressed to Gonzalo Garcia-Castro, Ramon Trial Fargas, 25-27, 08005 Barcelona. E-mail: \href{mailto:gonzalo.garciadecastro@upf.edu}{\nolinkurl{gonzalo.garciadecastro@upf.edu}}

}

\affiliation{\phantom{0}}

\abstract{
There is compelling evidence that bilinguals access their lexicon in a language non-selective way. For example, bilinguals produce and translated cognates (words whose translation in the other language is form-similar) faster than non-cognates, suggesting that the phonology of both languages interact during word production and comprehension (Costa et al., 2000; Christoffels et al., 2006). Previous literature on this effect has often relied on measures of overall form-similarity to categorise words into cognates andnon-cognates, such as the Levenshtein distance. These measures partially ignore potential sources of variability during lexical access like vowel vs.~consonant overlap, overlap in stressed syllables, onset overlap or the distance in features between both translations. In this study, we explored the impact of some of these variables on non-native word translation task: Spanish and English participants listened to non-native words (Catalan or Spanish) and were prompted to type their translation in their native language. Critically, participants where unfamiliar with the testing language, ensuring that they were not able to translate words based on previous knowledge on their meaning, leaving phonological information as the only cue participants were able to exploit to translate words to their native language correctly. We analysed the probability of correct translations, adjusting for the amount of vowel overlap, consonant overlap, overlap at stress position, overlap at onset, and distance in features between replaced phonemes, including the average frequency of phonological neighbours of the target words are a covariate.
}

\begin{document}
\maketitle

@@

Data collection took place from 04th June, 2020 to 28th June, 2020. We collected data from 105 participants (\(M_{age}\) = 21.79, \(SD_{age}\) = 2.43, \(Range_{age}\) = 18-33). 73 participants were British English native speakers living in United Kingdom (47 female), and 73 participants were Spanish native speakers living in Spain (27 female). Participants in UK were recruited via Prolific (5£ compensation) and SONA (compensation in academic credits). Participants in Spain were contacted via announcements in Faculties, and were compensated 5€ or an Amazon voucher for the same value. Participants were asked to complete the experiment in a quiet place with good internet connection. We excluded data from participants that a) self-rated their oral and/or written skills in a second or third language as higher than 4 in a 5-point scale (\emph{n} = 1), b) were diagnosed with a language (\emph{n} = 2)\footnote{We originally planned to exclude participants that reported any visual impairment that glasses would not correct This item was phrased as ``Do you have normal or corrected-to-normal VISION? (Yes/No)'' in English, and as ``¿Tienes problemas de VISIÓN que unas gafas o lentes de contacto NO corrijan? (Sí/No).'' However, the proportion of Spanish participants that reported visual impairment was implausibly large (\emph{n} = 6, 18.75\%). This is possibly due to these participants using glasses dayly and not having read the item until the end, where it is indicated that the use of glasses is considered as normal vision.}, or c) did not contribute more than 80\% of valid trials (\emph{n} = 9).

@@

The experiment was implemented online using Psychopy/Pavlovia (\textbf{peirce2019?}). Participants accessed the study from a link provided by Prolific or SONA and completed the experiment from a browser (Chrome or Mozilla). Participants were informed about the aims of the study and gave informed consent for participating. Then, participants answered a series of questions about their demographic status, their language background, and the set up they were using for completing the study. Third, participants completed the experimental task. Participants were informed that they would listen to a series of pre-recorded words in Catalan or Spanish (English participants) or only Catalan (Spanish participants). They were instructed to listen to each word, guess its meaning in English (English participants) or Spanish (Spanish participants), and type their answer as soon as possible. English participants were randomly assigned to the list of Catalan or Spanish trials. Participants in the Catalan list were presented with 86 trials, and participants in the Spanish list were presented with 103 trials.

Each trial started with a yellow fixation dot presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a ``\textgreater{}'' symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the RETURN key to start and new trial. We excluded trials where participants did not type an existing word in the correspondent language, or did not type anything at all. Trials where the response was mistyped by only one character were counted as correct, as long as the respond did not correspond to a distinct word. Participants contributed a total of 8077 valid trials (5235 in Catalan, 2842 in Spanish). The task took approximately 15 minutes to be completed.

@@

Participants listened to one audio file in each trial. This audio file corresponded to a word in Catalan (for Spanish speakers, and for English participants allocated in the Catalan condition) or Spanish (for English speakers allocated in the Spanish condition). The audio files were the same ones used in child experiments conducted in the Laboratori de Recerca en Infancia of Universitat Pompeu Fabra (Barcelona, Spain). These audio files were recorded by a proficient Catalan-Spanish female bilingual from the Metropolitan Area of Barcelona in a child-directed manner. Catalan and Spanish words were recorded at 44,100 Hz in separate files in the same session, and then de-noised using Audacity and normalised at peak intensity using Praat (\textbf{boersma2018?}). The average duration of the audios was NA (\emph{SD} = NA, \emph{Min} = \(\infty\), \emph{Max} = \(-\infty\)). The average lexical frequency of the words was 4.29 (\emph{SD} = 0.59, \emph{Min} = 2.87, \emph{Max} = 6.14). The average lexical frequency of the Catalan audios was 4.31 seconds (\emph{SD} = 0.59, \emph{Min} = 2.87, \emph{Max} = 5.68), and the average lexical frequency of the Spanish audios was 4.32 seconds (\emph{SD} = 0.56, \emph{Min} = 2.87, \emph{Max} = 5.68). Lexical frequencies in Catalan and Spanish were extracted from SUBTLEX-CAT (\textbf{boada2013?}) and SUBTLEX-ESP (\textbf{cuetos2011?}), respectively, and are expressed as Zipf scores (\textbf{van-heuven2012?}; \textbf{zipf1949?}).

@@

We divided participants in three groups: English natives tested in Catalan (ENG-CAT), English natives tested in Spanish (ENG-SPA), and Spanish natives tested in Catalan (SPA-CAT). We analysed the accuracy of responses to each item by calculating the proportion of correct responses across participants, resulting in one data point per item and group.

We fit several Bayesian linear models (\textbf{sorensen2015?}), each including one of the aforementioned variables as predictor, and the proportion of correct responses as output. Continuous predictors were standardised and categorical predictors were sum-coded (\textbf{schad2020?}). Missing predictor scores were imputed before fitting the models using multiple imputation (\textbf{vanbuuren2013?}). To test and account for cross-group differences, we included a random intercept for each group. We compared models using leave-one-out cross-validation (LOO, (\textbf{vehtari2016?})). More information about the models and model comparison can be found in Appendix X.

All analyses were performed in R environment (\textbf{rcoreteam2019?}). We used the \texttt{tidyverse} family of R packages to process data and to generate figures, the \texttt{mice} R package (\textbf{vanbuuren2013?}) for multiple imputation of missing data, and the \texttt{brms} R package (\textbf{burkner2017?}) to fit and compare models, which uses the Stan probabilistic programming language language (\textbf{carpenter2016?}).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The role of cognateness in non-native spoken word recognition},
  pdfkeywords={cognate, word recognition, translation, non-native,
spoken word recognition},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{The role of cognateness in non-native spoken word recognition}
\author{true \and true \and true \and true}
\date{}

\begin{document}
\maketitle
\begin{abstract}
There is compelling evidence that bilinguals access their lexicon in a
language non-selective way. For example, bilinguals produce and
translated cognates (words whose translation in the other language is
form-similar) faster than non-cognates, suggesting that the phonology of
both languages interact during word production and comprehension (Costa
et al., 2000; Christoffels et al., 2006). Previous literature on this
effect has often relied on measures of overall form-similarity to
categorise words into cognates andnon-cognates, such as the Levenshtein
distance. These measures partially ignore potential sources of
variability during lexical access like vowel vs.~consonant overlap,
overlap in stressed syllables, onset overlap or the distance in features
between both translations. In this study, we explored the impact of some
of these variables on non-native word translation task: Spanish and
English participants listened to non-native words (Catalan or Spanish)
and were prompted to type their translation in their native language.
Critically, participants where unfamiliar with the testing language,
ensuring that they were not able to translate words based on previous
knowledge on their meaning, leaving phonological information as the only
cue participants were able to exploit to translate words to their native
language correctly. We analysed the probability of correct translations,
adjusting for the amount of vowel overlap, consonant overlap, overlap at
stress position, overlap at onset, and distance in features between
replaced phonemes, including the average frequency of phonological
neighbours of the target words are a covariate.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{non-native-language-comprehension-is-difficult-but-cognateness-helps}{%
\subsection{Non-native language comprehension is difficult, but
cognateness
helps}\label{non-native-language-comprehension-is-difficult-but-cognateness-helps}}

Humans are able to recognise spoken words without much effort, even in
adverse conditions such as compressed speech (De Haan 1982) or the loss
of segmental information (Warren 1970). But even in the most ideal of
the situations, the processes engaged in word recognition do not occur
without uncertainty. The speech input activates multiple candidate
lexical entries based on their similarity with the signal, and for word
recognition to take place, the lexical system must select one of the
candidates. There is extensive literature about how the number of
candidates, their frequency, and their phonological and semantic
similarity with the target word impacts the dynamics of word
recognition. However, less is known about how these factors affect the
recognition of words in a non-native language.

Listening to speech in a non-native language is more costly than doing
so in the native language, even if one is proficient in the non-native
language. The source of this increased cognitive effort is likely to
originate at multiple levels. For instance, some sounds in the speech
signal do not correspond to any phoneme in the native language,
segmenting the speech signal relies on previous familiarity with word
forms or with the statistical regularities of the language, or word
order might be different in both languages (e.g., subject-object-verb
vs.~subject-verb-object.). The listener, however, is rarely completely
naïve to the language they are listening to. All languages share, to
some extent, similarities, frequently due to their typological
closeness, which can be exploited by listeners of a non-native language.

Lexical similarity, for example, provides useful information to
sequential language learners. Romance language like Spanish, Italian,
French or Catalan, share many similarities at the lexical level in the
form of cognates (Schepens, Dijkstra, and Grootjen 2012). Cognates are
cross-language synonyms whose form (e.g., phonology, orthography,
signature) is similar. The cause of this similarity is frequently
attributed to a shared etymological origin. This is the case of
\emph{puerta} and \emph{porta} (\emph{door} in Spanish and Catalan,
respectively)\footnote{Some form-similar cross-language synonyms are
  technically not cognates. For example, \emph{sun} and \emph{sol} (in
  Spanish), share their phonological onset, but their etymology points
  to different origins. We will use the word \emph{cognateness} to
  include all form-similar cross-language synonyms for simplicity. It is
  highly implausible that etymology plays a direct role on language
  perception if it is not via form-similarity, since it is not necessary
  for participants in psycholinguistic experiments to be aware of the
  etymology of the words they encounter in the tasks to be subject to
  the effect of form-similarity.}. There is evidence that cognates are
learnt more easily in L2 than non-cognates. De Groot and Keijzer (2000)
presented 40 Dutch natives 60 pairs of translation equivalents. In each
learning trial, two words were presented in a screen side-by-side: one
in Dutch and one pseudoword. Pseudowords were generated in such way that
were easily pronounceable and were phonotactically legal in Dutch. Word
pairs varied in their form-similarity (number of shared letters,
40-70\%). Participants were tested twice in the same task with one week
of difference. Participants' performance was better for cognates across
both testing sessions, suggesting that cognates were learnt and retained
more easily than non-cognates. Converging evidence was provided by Lotto
and De Groot (1998) in Dutch natives learning Italian words: cognates
were learnt more easily than non-cognates.

Cognates are not only learnt faster than non-cognates, but also
translated faster. Early accounts of bilingual lexical access described
two possible routes for translating a word from L1 to L2 or \emph{vice
versa}: (1) via direct links between L1-L2 links or (2) through
word-concept links. Both translation routes are available during
translation (Potter et al. 1984; Kroll, Curley, and others 1988). These
models assumed that the strength of the word-concept associations in
both L1 and L2 is a function of their frequency of use, in order to
account for the influence of both the lexical frequency of the presented
word and the target word in translation tasks on participant's
performance. Under the assumption that bilinguals use L1 more often,
Kroll and Stewart (1994) proposed the asymmetric hypothesis, predicting
stronger word-concept associations between L1 representations compared
to L2 representations. This would make L2-to-L1 translation more reliant
to the direct translation route (L1-L2 association) than L1-to-L2
translation. This prediction has three consequences: (1) L2-to-L1
translation is expected to be faster than L1-to-L2 translation (as it
does not rely on conceptual mediation during translation), and (2) since
word-to-word connections (like the L1-L2 one) occurs at the lexical
level (and not at the conceptual level), it should be sensitive to the
form-similarity between both words, and therefore to cognateness, and
(3) since low-proficiency bilinguals rely more on the direct route
during translation, and this route is sensitive to cognateness, their
performance during backward translation tasks should not only be better
than during forward translation, but also enhanced by the cognate status
of L1 and L2 representations.

To test these predictions, Degroot, Dannenburg, and Vanhell (1994) and
Groot (1992) asked 52 Dutch natives with high (yet non-native) English
proficiency to translate words from either Dutch to English (forward
translation, L1 to L2) or from English to Dutch (backward translation,
L2 to L1). In each trial, participants were presented visually with a
word in English or Dutch, and were asked to speak out loud its
translation in the other language as soon as possible. The authors
reported three main findings: (1) reaction times and accuracy were
roughly equivalent in across both forward and backward conditions
(although slightly faster in the former), (2) semantic variables, such
as imageability, were positively associated with participants'
performance more strongly in the forward translation conditional,
although the effect size was small, (3) cognates were translated equally
fast in both conditions, but non-cognates were translated faster during
backward translation than during forward translation, and (4) when
translating cognates participants' performance across both conditions
was less sensitive to semantic variables than when translating
non-cognates. These results prompted the authors to reject a hard
version Kroll and Stewart (1994)'s asymmetric hypothesis, and suggest
that both conceptually mediated and direct translation routes are active
during translation, but backward translation relies more strongly on
direct links between L1 and L2 representations, which makes it more
sensitive to cognateness than forward translation. When the authors
tested a group of bilinguals with higher proficiency in English, their
results pointed in the same direction.

The idealised lexicon considered by the models describe above only
considers word-word connections between translation equivalents. More
recent studies have highlighted the role of the rich network of
connections that given word establishes with other phonologically or
conceptually related words. Collins and Loftus (1975) introduced this
notion into their theory of semantic processing, and later Luce and
Pisoni (1998) operationalised this dimension in their Neighbourhood
Activation Model (NAM). In this model, lexical selection is mediated not
only by the lexical frequency of the target word, and the number of
activated candidates, but also by the frequency of the candidates. Luce
and Pisoni (1998) designed a lexical decision task in which English
native participants listened to a word across several conditions of
noise-to-signal ratio. The authors then used a computerised lexicon to
calculate the number of phonological neighbours around each of the
presented words, based on phonetic similarity matrices, and the average
lexical frequency of such neighbourhoods. Participants answered more
faster and more accurately to high frequency words. Words from high
density neighbourhoods were responded to more accurately, but more
slowly. The average lexical frequency of the neighbourhood was
associated to a decrease in accuracy. Low frequency words were responded
to more accurately in low density neighbours than in high density
neighbours. The authors concluded that, although both word lexical
frequency and neighbourhood density can, in principle, facilitate spoken
word recognition, this effect was modulated by the frequency of the
neighbourhood: when phonological neighbours are more frequent, lexical
selection is hindered (Goldinger, Luce, and Pisoni 1989; Luce, Pisoni,
and Goldinger 1990).

The role of neighbourhood lexical frequency on bilingual spoken word
recognition is unclear. Across several modalities, bilinguals'
recognition of words in a language is affected by form-similarity and
frequency of words in the other language. The parallel activation
hypothesis accounts for this phenomenon, suggesting that lexical access
is language-non selective. There is strong evidence supporting this
claim in both comprehension, production, and translation, and across
modalities. Much of this evidence stems from the fact that bilinguals'
production times are faster when naming cognates (i.e., words whose
translation in the other language is form-similar) than non-cognates.
Both phonological and semantic links between translations underlie
facilitatory effect of cognateness. The activation of form-similar words
in the non-target language that do not share meaning with the target
word, however, interferes with lexical selection. Disentangling the
effect of semantic and phonological links during spoken word recognition
is challenging when participants are bilingual.

This has important implications to non-native word recognition: is word
recognition facilitated by cognateness via translation? The answer is
not straightforward. Disentangling the effect of form- and conceptual
similarity in a bilingual sample is daunting, given that second language
learners are able to exploit \ldots{}

In the present study, we used a fully monolingual sample to disentangle
the effect of neighbourhood frequency and phonological similarity. We
asked participants to translate word in a language they had no
significant experience with. This made sure that participants were
unaware of any semantic relationship between the (non-native) words they
heard and their translation, and could only rely on their phonological
similarity to guess their translation. We tested the role of the amount
of phonological similarity between the presented and the target words,
the lexical frequency of the target word, and the average lexical
frequency of the target word's phonological neighbourhood.

@@

Data collection took place from June 04th, 2020 to June 28th, 2020. We
collected data from 104 participants (\emph{Mean} = 21.79 years,
\emph{SD} = 2.43, \emph{Min} = 18, \emph{Max} = 33). 72 participants
were British English native speakers living in United Kingdom (46), and
32 participants were Spanish native speakers living in Spain (27
female). Participants in UK were recruited via Prolific (5£
compensation) and SONA (compensation in academic credits). Participants
in Spain were contacted via announcements in Faculties, and were
compensated 5€ or an Amazon voucher for the same value. Participants
were asked to complete the experiment in a quiet place with good
internet connection. We excluded data from participants that a)
self-rated their oral and/or written skills in a second or third
language as higher than 4 in a 5-point scale (\emph{n} = 1), b) were
diagnosed with a language (\emph{n} = 2) , or c) did not contribute more
than 80\% of valid trials (\emph{n} = 9).

@@

The experiment was implemented online using Psychopy/Pavlovia (Peirce et
al. 2019). Participants accessed the study from a link provided by
Prolific or SONA and completed the experiment from a browser (Chrome or
Mozilla). Participants were informed about the aims of the study and
gave informed consent for participating. This experiment was Then,
participants answered a series of questions about their demographic
status, their language background, and the set up they were using for
completing the study. Third, participants completed the experimental
task. Participants were informed that they would listen to a series of
pre-recorded words in Catalan or Spanish (English participants) or only
Catalan (Spanish participants). They were instructed to listen to each
word, guess its meaning in English (English participants) or Spanish
(Spanish participants), and type their answer as soon as possible.
English participants were randomly assigned to the list of Catalan or
Spanish trials. Participants in the Catalan list were presented with 86
trials, and participants in the Spanish list were presented with 103
trials.

Each trial started with a yellow fixation point presented during one
second on the centre of the screen over a black background. After one
second, the audio started playing while the dot remained being displayed
until the audio ended. Upon the offset of the fixation point and audio,
participants were prompted to write their answer by a ``\textgreater{}''
symbol. Typed letters were displayed in the screen in real time to
provide visual feed-back to participants. Participants were allowed to
correct their answer. Then, participants pressed the RETURN key to start
and new trial. We excluded trials where participants did not type an
existing word in the correspondent language, or did not type anything at
all. Trials where the response was mistyped by only one character were
counted as correct, as long as the respond did not correspond to a
distinct word. Participants contributed a total of 8077 valid trials
(5235 in Catalan, 2842 in Spanish). The task took approximately 15
minutes to be completed.

@@

Participants listened to one audio file in each trial. This audio file
corresponded to a word in Catalan (for Spanish speakers, and for English
participants allocated in the Catalan condition) or Spanish (for English
speakers allocated in the Spanish condition). The audio files were the
same ones used in child experiments conducted in the Laboratori de
Recerca en Infància of Universitat Pompeu Fabra (Barcelona, Spain).
These audio files were recorded by a proficient Catalan-Spanish female
bilingual from the Metropolitan Area of Barcelona in a child-directed
manner. Catalan and Spanish words were recorded at 44,100 Hz in separate
files in the same session, and then de-noised using Audacity and
normalised at peak intensity using Praat (Broersma and Weenink 2021).
The average duration of the audios was 1.20 (\emph{SD} = 0.18,
\emph{Min} = 0.78, \emph{Max} = 1.58). The average duration of the
Catalan audios was 1.23 seconds (\emph{SD} = 0.19, \emph{Min} = 0.80,
\emph{Max} = 1.58), and the average duration of the Spanish audios was
1.16 seconds (\emph{SD} = 0.15, \emph{Min} = 0.78, \emph{Max} = 1.53).
Table 1 summarises the lexical frequency, phonological neighbourhood
density and phonological overlap of the words included in the Catalan
and the Spanish lists.

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{lrrrrrrrrrr}
\toprule
 & \multicolumn{2}{c}{\textbf{Frequency (per million)}} & \multicolumn{2}{c}{\textbf{Frequency (Zipf score)}} & \multicolumn{2}{c}{\emph{\textbf{PTHN}}} & \multicolumn{2}{c}{\textbf{Consonant similarity}} & \multicolumn{2}{c}{\textbf{Vowel similarity}} \\ 
 \cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9} \cmidrule(lr){10-11}
 & Mean & \emph{SD} & Mean & \emph{SD} & Mean & \emph{SD} & Mean & \emph{SD} & Mean & \emph{SD} \\ 
\midrule
ENG-CAT & $46.80$ & $77.23$ & $4.31$ & $0.59$ & $3.26$ & $3.74$ & $0.34$ & $0.17$ & $0.17$ & $0.11$ \\ 
ENG-SPA & $43.91$ & $67.53$ & $4.32$ & $0.56$ & $3.52$ & $4.10$ & $0.34$ & $0.15$ & $0.14$ & $0.10$ \\ 
SPA-CAT & $53.32$ & $156.19$ & $4.23$ & $0.62$ & $1.62$ & $2.62$ & $0.42$ & $0.15$ & $0.27$ & $0.12$ \\ 
 \bottomrule
\end{longtable}

@@

We modelled the probability of participants guessing the correct
translation of each input word using a generalised multilevel Bayesian
regression model with a Bernoulli logit link distribution. We first
fitted a base model (Model 0) that only included the lexical frequency
(\texttt{frequency}) of the target word as a fixed effect, with a random
intercept per participant. Second, we extended the model to include
\texttt{pthn} as a fixed effect, with a random slope by participant
(Model 1). Third, we added the fixed effect \texttt{consonant\_ratio}
and the \texttt{pthn:consonant\_ratio}, and random slopes for both
effects by participant (Model 2). Third, we added the fixed effect
\texttt{vowel\_ratio} and the \texttt{pthn:vowel\_ratio}, and random
slopes for both effects by participant (Model 3). Finally, we fit a
model that included the two-way interactions
\texttt{pthn:consonant\_ratio} and \texttt{pthn:vowel\_ratio}, an their
random slopes by participant (Model 4). Continuous predictors were
standardised and categorical predictors were sum-coded (Schad et al.
2020). Model 4 can be formally expressed as:

\$\$ \begin{align*}

&\textbf{Likelihood}  \\
y_{i} \sim& Bernoulli(p_{i}) && \text{[probability of correct translation]} \\ \\

&\textbf{Parameters}  \\

logit(p_{i}) = ~ &  \beta_{0[p,w]} ~ +  && \text{[linear model]}\\
& \beta_{1[p]} ~ Frequency_{i} ~ + \\
& \beta_{2[p]} ~ PTHN_i ~ + \\
& \beta_{3[p]} ~ Similarity_i ~ + \\
& \beta_{4[p]} ~ (PTHN_i \times Similarity_i) \\ \\

\beta_{0-6[p,w]} \sim& ~  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p ~\text{in 1, ..., } P ~\text{and  word } w ~\text{in 1, ..., } W && \text{[participant- and word-level intercepts]} \\
\beta_{1-6[p]} \sim& ~  \mathcal{N}(\mu_{\beta_{j}}, \sigma_{\beta_{j}}) \text{, for participant } p ~\text{in 1, ..., } P
&& \text{[participant-level coefficients]} \\ \\

&\textbf{Prior}  \\
